Preliminary Data Analysis

### 2023-08-17 Preliminary Analyses/Preliminary Data Analysis.Rmd
Full prelim analyses completed. For 2024 MSC proposal

library(tidyverse)
library(rio)
library(gt)

@@

analysis_data %>%
  dplyr::select(c(id:pretest, speaker)) %>%
  tidyr::pivot_longer(cols = posttest:pretest,
                      names_to = "phase",
                      values_to = "intel") %>%
  dplyr::mutate(phase = as.factor(phase),
                phase = fct_relevel(phase, "pretest", "posttest")) %>%
  ggplot() +
  aes(x = phase,
      y = intel,
      group = id,
      color = speaker) +
  geom_line(alpha = 0.3) +
  stat_summary(aes(group = speaker),
               fun.y = mean, geom = "line", linewidth = 2) +
  facet_wrap("speaker") +
  scale_color_manual(values = my_pal) +
  labs(x = "Phase", y = "Intelligibility") +
  theme_bw() +
  theme(legend.position = "none")

@@

>>>>

library(tidyverse)
library(rio)
library(gt)
library(reshape2) #install.packages("reshape2")
library(ggcorrplot) #install.packages("ggcorrplot")
library(patchwork) #install.packages("patchwork")

@@

pre_post <- analysis_data %>%
  dplyr::select(c(id:pretest, speaker)) %>%
  tidyr::pivot_longer(cols = posttest:pretest,
                      names_to = "phase",
                      values_to = "intel") %>%
  dplyr::mutate(phase = as.factor(phase),
                phase = fct_relevel(phase, "pretest", "posttest")) %>%
  ggplot() +
  aes(x = phase,
      y = intel,
      group = id,
      color = speaker) +
  geom_line(alpha = 0.3) +
  stat_summary(aes(group = speaker),
               fun.y = mean, geom = "line", linewidth = 2) +
  facet_wrap("speaker") +
  scale_color_manual(values = my_pal) +
  labs(x = "Phase", y = "Intelligibility") +
  theme_bw() +
  theme(legend.position = "none", aspect.ratio = 1)

pre_post

@@

## Cognitive Subtests

In this section, we will now look at the listeners' scores on the subtests from the NIH Toolbox Cognitive Battery. The assessments that were administered to the participants are listed below. In this analysis, we are using the age-corrected standard scores.

-   List Sort Working Memory Test (list_sort_c)

-   Flanker Inhibitory Control and Attention Test (flanker_c)

-   Dimensional Change Card Sort Test (card_sort_c)

    -   We also refer to this as cognitive flexibility

-   Picture Vocabulary Test (vocab_c)

-   Pattern Comparison Processing Speed Test (pattern_c)

-   Picture Sequence Memory Test (pic_seq_c)

The dataset also contains the words-in--noise thresholds for the participants' right (win_r) and left (win_l). We determined the participant's best threshold (e.g. the ear with the lower threshold score) and created a new variable called best_win. To see the full data processing steps, click "show code".

```{r}

# Determining ear with the lowest words-in-noise threshold (best_win)

analysis_data <- analysis_data %>%
  dplyr::mutate(best_win = ifelse(win_r < win_l, win_r, win_l)) %>%
  dplyr::select(!c(win_r, win_l, fluid_cog_c))

### Overall Cognitive Scores

The descriptive statistics for performance on each of the cognitive subtests are

cog_des <- analysis_data %>%
  dplyr::select(!c(posttest:speaker)) %>%
  tidyr::pivot_longer(cols = list_sort_c:best_win,
                      names_to = "cog",
                      values_to = "score") %>%
  dplyr::group_by(cog) %>%
  dplyr::summarize(Mean = mean(score, na.rm = T), SD = sd(score, na.rm = T)) %>%
  gt::gt() %>%
  fmt_number(columns = "Mean":"SD", decimals = 2) %>%
  cols_label(cog = " ")

cog_des

cog_des %>%
  gt::gtsave("Overall_cog.html", path = "Tables")

### Cognitive Subtest Scores per Training Group

We are now comparing the mean and standard deviation in these cognitive scores for listeners who were trained with AM1 vs PDM10.

cog_by_speaker <- analysis_data %>%
  dplyr::select(!c(posttest:improv)) %>%
  tidyr::pivot_longer(cols = list_sort_c:best_win,
                      names_to = "cog",
                      values_to = "score") %>%
  dplyr::group_by(cog, speaker) %>%
  dplyr::summarize(Mean = mean(score, na.rm = T), SD = sd(score, na.rm = T)) %>%
  dplyr::ungroup() %>%
  tidyr::pivot_wider(names_from = speaker,
                     values_from = c(Mean:SD)) %>%
  dplyr::relocate(SD_AM1, .before = Mean_PDM10) %>%
  gt::gt() %>%
  fmt_number(columns = "Mean_AM1":"SD_PDM10", decimals = 2) %>%
  tab_spanner(label = "AM1",
              columns = c(Mean_AM1, SD_AM1)) %>%
  tab_spanner(label = "PDM10",
              columns = c(Mean_PDM10, SD_PDM10)) %>%
  cols_label(cog = " ",
             Mean_AM1 = "Mean",
             SD_AM1 = "SD",
             Mean_PDM10 = "Mean",
             SD_PDM10 = "SD")

cog_by_speaker

cog_by_speaker %>%
  gt::gtsave("cog_by_speaker.html", path = "Tables")

# Correlations

In this study, we are particularly interested in the relationships between these cognitive subtests and initial intelligibility (pretest) and intelligibility difference between pretest and posttest (improv). In this preliminary analysis, we first started by calculating the correlations between these measures for 1) listeners trained with AM1, and 2) listeners trained with PDM10.

## Correlation Matrix for Listeners assigned to AM1

AM1_corr <- analysis_data %>%
  dplyr::filter(speaker == "AM1") %>%
  dplyr::select(c(pretest:improv, list_sort_c:best_win)) %>%
  dplyr::rename("Initial Intelligibility" = pretest,
                "Intelligibility Improvement" = improv,
                "Working Memory" = list_sort_c,
                "Inhibitory Control" = flanker_c,
                "Cognitive Flexibility" = card_sort_c,
                "Receptive Vocabulary" = vocab_c,
                "Processing Speed" = pattern_c,
                "Speech in Noise Threshold" = best_win,
                "Picture Sequence Memory" = pic_seq_c)

AM1_corr <- round(cor(AM1_corr, use = "pairwise.complete.obs"), 2)

get_lower_tri<-function(AM1_corr){
    AM1_corr[upper.tri(AM1_corr)]<- NA
    return(AM1_corr)
  }

AM1_lower_cor <- get_lower_tri(AM1_corr)

AM1_lower_cor <- melt(AM1_lower_cor, na.rm = T)

AM1 <- AM1_lower_cor %>%
  ggplot() +
  aes(x = Var1,
      y = Var2,
      fill = value) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = my_pal[1], high = my_pal[1], mid = "white",
                      midpoint = 0, limit = c(-1,1), space = "Lab",
                      name = "Pearson\nCorrelation") +
  geom_text(aes(Var1, Var2, label = value), color = "black", size = 3) +
  labs(x = " ",
       y = " ") +
  ggtitle('AM1') +
  theme_bw() +
  theme(axis.text.x = element_text(color = "black", angle = 45, hjust = 1),
        aspect.ratio = 1)

AM1

ggsave("Plots/AM1_corr.png", plot = last_plot())

## Correlation Martix for Listeners Assigned to PDM10

PDM10_corr <- analysis_data %>%
  dplyr::filter(speaker == "PDM10") %>%
  dplyr::select(c(pretest:improv, list_sort_c:best_win)) %>%
  dplyr::rename("Initial Intelligibility" = pretest,
                "Intelligibility Improvement" = improv,
                "Working Memory" = list_sort_c,
                "Inhibitory Control" = flanker_c,
                "Cognitive Flexibility" = card_sort_c,
                "Receptive Vocabulary" = vocab_c,
                "Processing Speed" = pattern_c,
                "Speech in Noise Threshold" = best_win,
                "Picture Sequence Memory" = pic_seq_c)

PDM10_corr <- round(cor(PDM10_corr, use = "pairwise.complete.obs"), 2)

get_lower_tri<-function(PDM10_corr){
    PDM10_corr[upper.tri(PDM10_corr)]<- NA
    return(PDM10_corr)
  }

PDM10_lower_cor <- get_lower_tri(PDM10_corr)

PDM10_lower_cor <- melt(PDM10_lower_cor, na.rm = T)

PDM10 <- PDM10_lower_cor %>%
  ggplot() +
  aes(x = Var1,
      y = Var2,
      fill = value) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = my_pal[2], high = my_pal[2], mid = "white",
                      midpoint = 0, limit = c(-1,1), space = "Lab",
                      name = "Pearson\nCorrelation") +
  geom_text(aes(Var1, Var2, label = value), color = "black", size = 3) +
  labs(x = " ",
       y = " ") +
  ggtitle('PDM10') +
  theme_bw() +
  theme(axis.text.x = element_text(color = "black", angle = 45, hjust = 1),
        aspect.ratio = 1)

PDM10

ggsave("Plots/PDM10_corr.png", plot = last_plot())

```{r, echo = F, eval = F}

figure_1 <- (AM1 | PDM10) /
  pre_post

figure_1

ggsave("Plots/Full_figure.png", plot = last_plot(), width = 15, height = 15, units = "in")
